Java Sorting Algorithms Comparison

### 2015-02-22 Programming Project B/paper.tex
added heapsort pseudocode to paper
gathered more testing results

A merge sort can be implemented in a variety of languages using the pseudocode provided in Algorithm 2.
\newline
\textbf{Merge Sort Pre-Condition}: A is a non-empty array of a comparable data with a natural order.
\newline
\textbf{Merge Sort Post-Condition}: A' is a permutation of A (containing all the same elements) in strictly non-decreasing order.
\newline
\textbf{Merge Pre-Condition}: Left and right are both non-empty arrays of a comparable data type with a natural order in strictly non-decreasing order.
\newline
\textbf{Merge Post-Condition}: Combined has all the elements of both left and right in strictly non-decreasing order.
\begin{algorithm}
\caption {\textsc{Merge-Sort}(A)}
\label{algo:mergesort}
\begin{algorithmic}[1]
\Procedure{Merge-Sort}{A}
\If{$A.length < 2$}
\State{\Return{$A$}}
\EndIf
\State{$mid = A.length / 2$}
\For{$i = 1$ upto $mid$}
	\State{$left[left.length] = A[i]$}
\EndFor
\For{$i = mid$ upto $A.length$}
	\State{$right[right.length] = A[i]$}
\EndFor
\State{$left =$ Merge-Sort($left$)}
\State{$right =$ Merge-Sort($right$)}
\State{$A =$ Merge($left$, $right$)}
\State{\Return{$A$}}
\EndProcedure
\newline
\Procedure{Merge}{left, right}
\State{$var$ $i = 0$}
\State{$var$ $y = 0$}
\State{$var$ $x = 0$}
\While{$left.length$ $!=$ $i$ and $right.length$ $!=$ $y$}
	\If{$left[i] < right[y]$}
		\State{$combined[x] = left[i]$}
		\State{$i = i + 1$}
		\State{$x = x + 1$}
	\EndIf
	\If{$right[y] < left[i]$}
		\State{$combned[x] = right[y]$}
		\State{$y = y + 1$}
		\State{$x = x + 1$}
	\EndIf
\EndWhile
\While{$left.length$ $!=$ $i$}
	\State{$combined[x] = left[i]$}
	\State{$i = i + 1$}
	\State{$x = x + 1$}
\EndWhile
\While{$right.length$ $!=$ $y$}
	\State{$combined[x] = right[y]$}
	\State{$y = y + 1$}
	\State{$x = x + 1$}
\EndWhile
\State{\Return{$combined$}}
\EndProcedure
\end{algorithmic}
\end{algorithm}
\newline
%% END PROCEDURE

%% BEGIN TESTING
\section{Testing}
\subsection{Testing Plan and Results}

\subsection{Problems Encountered}

>>>>

A merge sort can be implemented in a variety of languages using the pseudocode provided in Algorithm 2.
\newline
\textbf{Merge Sort Pre-Condition}: A is a non-empty array of a comparable data with a natural order.
\newline
\textbf{Merge Sort Post-Condition}: A' is a permutation of A (containing all the same elements) in strictly non-decreasing order.
\newline
\textbf{Merge Pre-Condition}: Left and right are both non-empty arrays of a comparable data type with a natural order in strictly non-decreasing order.
\newline
\textbf{Merge Post-Condition}: Combined has all the elements of both left and right in strictly non-decreasing order.
\begin{algorithm}
\caption {\textsc{Merge-Sort}(A)}
\label{algo:mergesort}
\begin{algorithmic}[1]
\Procedure{Merge-Sort}{A}
\If{$A.length < 2$}
\State{\Return{$A$}}
\EndIf
\State{$mid = A.length / 2$}
\For{$i = 1$ upto $mid$}
	\State{$left[left.length] = A[i]$}
\EndFor
\For{$i = mid$ upto $A.length$}
	\State{$right[right.length] = A[i]$}
\EndFor
\State{$left =$ Merge-Sort($left$)}
\State{$right =$ Merge-Sort($right$)}
\State{$A =$ Merge($left$, $right$)}
\State{\Return{$A$}}
\EndProcedure
\newline
\Procedure{Merge}{left, right}
\State{$var$ $i = 0$}
\State{$var$ $y = 0$}
\State{$var$ $x = 0$}
\While{$left.length$ $!=$ $i$ and $right.length$ $!=$ $y$}
	\If{$left[i] < right[y]$}
		\State{$combined[x] = left[i]$}
		\State{$i = i + 1$}
		\State{$x = x + 1$}
	\EndIf
	\If{$right[y] < left[i]$}
		\State{$combned[x] = right[y]$}
		\State{$y = y + 1$}
		\State{$x = x + 1$}
	\EndIf
\EndWhile
\While{$left.length$ $!=$ $i$}
	\State{$combined[x] = left[i]$}
	\State{$i = i + 1$}
	\State{$x = x + 1$}
\EndWhile
\While{$right.length$ $!=$ $y$}
	\State{$combined[x] = right[y]$}
	\State{$y = y + 1$}
	\State{$x = x + 1$}
\EndWhile
\State{\Return{$combined$}}
\EndProcedure
\end{algorithmic}
\end{algorithm}
\newline
%% TODO: Add pre and post conditions here
A heap sort can be implemented in a variety of languages using the pseudocode below in Algorithm 3.
\newline
\begin{algorithm}
\caption {\textsc{Heap-Sort}(A)}
\label{algo:heapsort}
\begin{algorithmic}[1]
\Procedure{Heapify}{A, i, total}
\State{$var$ $left = i * 2$}
\State{$var$ $right = left + 1$}
\State{$var$ $iPrime = i$}
\If{$left <= total$ and $A[left] > A[i]$}
	\State{$i = left$}
\EndIf
\If{$right <= total$ and $A[right] > A[i]$}
	\State{$i = right$}
\EndIf
\If{$i != iPrime$}
	\State{$var$ $temp = A[iPrime]$}
	\State{$A[iPrime] = A[i]$}
	\State{$A[i] = tmp$}
	\State{$A = heapify(A, i, total)$}
\EndIf
\State{\Return{$A$}}
\EndProcedure
\Procedure{Heap-Sort}{A}
\State{$var$ $size = A.length$}
\For{$var$ $i = size / 2$ downto $1$}
	\State{$A = heapify(A, i, size$}
\EndFor
\For{$var$ $i = size$ downto $1$}
	\State{$var$ $tmp = A[1]$}
	\State{$A[0] = A[i]$}
	\State{$A[i] = tmp$}
	\State{$size = size - 1$}
	\State{$A = heapify(A, 1, size)$}
\EndFor
\State{\Return{$A$}}
\EndProcedure
\end{algorithmic}
\end{algorithm}
%% END PROCEDURE

%% BEGIN TESTING
\section{Testing}
\subsection{Testing Plan and Results}
All arrays used in testing are Java ArrayList<Integer> unless otherwise spcified. All times are recorded in milliseconds using a stopwatch class borrowed from Robert Sedgewock and Kevin Wayne %%TODO: Cite this shit.
It is important to note that the soptwatch class used takes the elapsed real-time between the start of the sort algorithm and the end of the sort algorithim as opposed to taking the elapsed processor-time becasue these tests were run on a multi-core computer.
In the table below, A denotes Array. Times in the table below are given as averages out of 10 trials.
\newline
\captionof{table}{Insertion Sort Test Results}
\begin{center}
\begin{tabular}{|l|l|l|l|}
\hline Tested Input & Expected Results & Actual Results & Time \\
\hline Empty A & Empty A & Empty A & 0.0003 \\
\hline A of 1000 Strings & Sorted A 1000 Strings & Sorted A 1000 Strings & 0.021 \\
\hline A 1 Element & Original A & Original A & 0.0003 \\
\hline A 10 Elements & Sorted A 10 Elements & Sorted A 10 Elements & 0.0005 \\
\hline A 100 Elements & Sorted A 100 Elements & Sorted A 100 Elements & 0.0021 \\
\hline A 1000 Elements & Sorted A 1000 Elements & Sorted A 1000 Elements & 0.019 \\
\hline A 10000 Elements & Sorted A 10000 Elements & Sorted A 10000 Elements & 0.129 \\
\hline A 100000 Elements & Sorted A 100000 Elements & Sorted A 100000 Elements & 6.4923 \\
\hline A 1000000 Elements & Sorted A 1000000 Elements & Sorted A 1000000 Elements & 2135.5007 \\
\hline A 10000000 Elements & Sorted A 10000000 Elements & OS Crash & N/A \\
\hline A 1000 Identical Elements & Original Array & Original Array & 0.0052 \\
\hline
\end{tabular}
\end{center}
\newline
\captionof{table}{Merge Sort Test Results}
\begin{center}
\begin{tabular}{|l|l|l|l|}
\hline Tested Input & Expected Results & Actual Results & Time \\
\hline Empty A & Empty A & Empty A & 0.0002 \\
\hline A 1 Element & Original A & Original A & 0.0001 \\
\hline A 10 Elements & Sorted A 10 Elements & Sorted A 10 Elements & 0.0004 \\
\hline A 100 Elements & Sorted A 100 Elements & Sorted A 100 Elements & 0.0028 \\
\hline A 1000 Elements & Sorted A 1000 Elements & Sorted A 1000 Elements & 0.0294 \\
\hline A 10000 Elements & Sorted A 10000 Elements & Sorted A 10000 Elements & 0.2098 \\
\hline A 100000 Elements & Sorted A 100000 Elements & Sorted A 100000 Elements & 2.1904 \\
\hline A 1000000 Elements & Sorted A 1000000 Elements & Sorted A 1000000 Elements & 22.9314 \\
\hline 
\hline
\end{tabular}
\end{center}
\subsection{Problems Encountered}
